---
title: "Week 5 Practice Problems"
author: 
  name: Alex Kaizer
  roles: "Instructor"
  affiliation: University of Colorado-Anschutz Medical Campus
toc: true
toc_float: true
toc-location: left
format:
  html:
    code-fold: show
    code-overflow: wrap 
    code-tools: true
---

```{r, echo=F, message=F, warning=F}
library(kableExtra)
library(dplyr)
```

This page includes optional practice problems, many of which are structured to assist you on the homework with [Solutions provided on a separate page](/labs/prac5s/index.qmd). Data sets, if needed, are provided on the BIOS 6618 Canvas page for students registered for the course.

This week's extra practice exercises are focusing on implementing bootstrap resampling and permutation testing to evaluate the odds ratio of an estimate.

# Data Background

Complete the following exercises to conduct a bootstrap and a permutation test for a data set we first used in last week's lab.

The following code can load the `Surgery_Timing.csv` file into R. The surgery time data is based on a retrospective observational study of 32,001 elective general surgical patients, but we will subset to arthroplasty knee procedures. We will create a new variable to specify morning vs. afternoon surgery time as our "exposure" and will examine in-hospital complication rate as our "outcome" of interest.

```{r, cache=T, class.source = 'fold-show'}
dat1 <- read.csv('../../.data/Surgery_Timing.csv')
dat1s <- dat1[which(dat1$ahrq_ccs=='Arthroplasty knee'),]
dat1s$AK_morning <- dat1s$hour < 12 # create new variable for morning observation, I use the "AK_" prefix to indicate variables I created
```

With this information we can calculate the odds ratio in a few ways (manually, with a function, etc.):

```{r, message=FALSE, class.source = 'fold-show', warning=F}
# calculate the OR from epi.2by2
library(epiR)
tab1 <- table(morning = factor(dat1s$AK_morning, levels=c(TRUE,FALSE)), complication = factor(dat1s$complication, levels=c(1,0)) )
epi.2by2(tab1)

# calculate the OR by hand where OR = ad/bc
a <- sum( dat1s$AK_morning==T & dat1s$complication==1 )
b <- sum( dat1s$AK_morning==T & dat1s$complication==0 )
c <- sum( dat1s$AK_morning==F & dat1s$complication==1 )
d <- sum( dat1s$AK_morning==F & dat1s$complication==0 )
obs_or <- (a*d)/(b*c)
obs_or
```

We see that in both approaches we arrive at an estimated odds ratio of 0.844, with a 95% CI from `epi.2by2` of (0.63, 1.13).

## Exercise 1: Bootstrap Confidence Intervals

Estimate the 95% normal percentile and bootstrap percentile confidence intervals with 10,000 bootstrap samples to describe the variability of our estimate and:

**a.** compare the resulting CIs to the estimate from `epi.2by2`

**b.** evaluate if the normal percentile CI has acceptable coverage

**c.** evaluate if the bootstrap percentile CI has acceptable accuracy


## Exercise 2: Permutation Test P-value

Implement a permutation test with 10,000 resamples to estimate a p-value for if our observed OR is significantly different from its null value for:

**a.** a two-sided p-value.

**b.** a one-sided p-value where we hypothesize that mornings have a lower odds of complications compared to the afternoon.




