---
title: "Mediation Modeling Example with Bootstrap CI Estimate"
author: 
  name: Alex Kaizer
  roles: "Instructor"
  affiliation: University of Colorado-Anschutz Medical Campus
toc: true
toc_float: true
toc-location: left
format:
  html:
    code-fold: show
    code-overflow: wrap 
    code-tools: true
---

```{r, echo=F, message=F, warning=F}
library(kableExtra)
library(dplyr)
```

This page is part of the University of Colorado-Anschutz Medical Campus' [BIOS 6618 Recitation](/recitation/index.qmd) collection. To view other questions, you can view the [BIOS 6618 Recitation](/recitation/index.qmd) collection page or use the search bar to look for keywords.

# Mediation Example Problem

We will use linear regression to examine the relationship between the outcome of IQ test score (`iq`) with lead exposure (`expose`) and years lived at the current residence (`resdur`) as a potential mediator to answer the following exercises:

## Three Fundamental Models of Mediation

Fit the three fundamental models of mediation analysis and fill in the following DAG:

```{tikz, tikz-med, fig.ext = 'png', cache=TRUE, echo=F, out.width="70%"}
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=4cm, auto,>=latex, thick, scale = 0.5]
\node (X) [align=center] {Exposure \\ ($X$)};
\node (M) [align=center,above of=X,xshift=2cm,yshift=-2cm] {Duration at Residence \\ ($M$)};
\node (Y) [align=center,right of=X] {IQ Score \\ ($Y$)};
\draw[->] (X) to node [above left] {$\hat{\gamma}_{X}=0.17$} (M);
\draw[->] (M) to node [above right] {$\hat{\beta}_{M}=-0.80$} (Y);
\draw[->] (X) to node [align=center,swap] {$\hat{\beta}_{adj}=-7.64$ \\ $\hat{\beta}_{crude}=-7.77$} (Y);
\end{tikzpicture}
```

**Solution:**
```{r, echo=T}
lead <- read.csv('../../.data/lead2.csv') # read in data

ex2_crude <- lm(iq ~ expose, data=lead) # fit crude model
ex2_adjusted <- lm(iq ~ expose + resdur, data=lead) # fit adjusted model
ex2_covariate <- lm(resdur ~ expose, data=lead) # fit covariate model

coef(ex2_crude)
coef(ex2_adjusted)
coef(ex2_covariate)
```



## Proportion Mediated by Duration at Residence

What is the proportion/percent mediated by duration at residence?

**Solution:**

The proportion mediated is $\frac{\text{indirect effect}}{\text{total effect}} = \frac{\hat{\beta}_{crude} - \hat{\beta}_{adj}}{\hat{\beta}_{crude}} = \frac{-7.77 - (-7.64)}{-7.77} = 0.0167$. This corresponds to the percent mediated of $0.0167 \times 100 = 1.67\%$.


## Confidence Interval Around Proportion Mediated and p-Value

What is the 95% CI and corresponding p-value for the proportion/percent mediated by duration at residence using the normal approximation to estimate the standard error (i.e., Sobel's test)?

**Solution:**

We first need to calculate the standard error for our indirect effect:
$$ SE(\hat{\beta}_{crude} - \hat{\beta}_{adj}) = \sqrt{{\hat{\gamma}}_X^2\left(SE\left({\hat{\beta}}_M\right)\right)^2+{\hat{\beta}}_M^2\left(SE\left({\hat{\gamma}}_X\right)\right)^2}  $$

We can find this information in our coefficient tables:

```{r, echo=T}
summary(ex2_adjusted)$coefficient
summary(ex2_covariate)$coefficient
```

$$ SE(\hat{\beta}_{crude} - \hat{\beta}_{adj}) = \sqrt{(0.17)^2(0.40)^2 + (-0.80)^2 (0.65)^2} = \sqrt{0.275} = 0.524 $$

Our 95% CI for the indirect effect is

$$ -0.13 \pm 1.96 \times 0.524 = (-1.16, 0.90)  $$

This corresponds to a 95% CI for our proportion mediated of:

$$ \left(\frac{-1.16}{-7.77}, \frac{0.90}{-7.77} \right) = (-0.1158, 0.1493) $$
The 95% CI may suggest an inconsistent mediation model since our confidence interval includes negative values or indicates an insignificant result, which we can verify by calculating the p-value.

Our p-value is calculated from referencing the standard normal distribution (per Sobel's test):

$$ Z = \frac{\text{indirect effect}}{SE(\hat{\beta}_{crude} - \hat{\beta}_{adj})} = \frac{-0.13}{0.524} = -0.248 $$

This corresponds to a p-value of `2 * pnorm(-0.248)`=`r round(2*pnorm(-0.248),3)`, which is not significant.


## Bootstrap Percentile CI for Proportion Mediated

Calculate the *95% bootstrap percentile confidence interval* for the proportion/percent mediated. How does it compare with your estimate from 2c?

**Solution:**

For our bootstrap, we will resample from the overall sample and implement 1,000 bootstrap resamples:

```{r, echo=T}
# set seed for reproducibility
set.seed(1631)

# implement bootstrap
B <- 1000
boot <- numeric(B)

for(i in 1:B){
  boot_dat <- lead[sample(1:nrow(lead), size=nrow(lead), replace=T),]
  boot_crude <- lm(iq ~ expose, data=boot_dat)
  boot_adjusted <- lm(iq ~ expose + resdur, data=boot_dat)
  boot[i] <- (boot_crude$coef['expose'] - boot_adjusted$coef['expose']) / boot_crude$coef['expose']
}

# 95% bootstrap percentile interval
quantile(boot, c(0.025,0.975))

# accuracy to evaluate performance of bootstrap percentile interval
bias_boot <- 0.0167 - mean(boot)
se_boot <- sd(boot)
abs(bias_boot)/se_boot
```

Our 95% bootstrap percentile confidence interval is (-0.24, 0.22) for the proportion mediated.This is wider than our estimated confidence interval based on the normal approximation for Sobel's test, but still reflects our insignificant result since the interval includes 0 (which may also suggest an inconsistent mediation model). The accuracy of our bootstrap percentile interval is acceptable given the bias/SE is less than 0.10.

