---
title: "Population Moments"
author: 
  name: Alex Kaizer
  roles: "Instructor"
  affiliation: University of Colorado-Anschutz Medical Campus
toc: true
toc_float: true
toc-location: left
format:
  html:
    code-fold: show
    code-overflow: wrap 
    code-tools: true
---

```{r, echo=F, message=F, warning=F}
library(kableExtra)
library(dplyr)
```

This page is part of the University of Colorado-Anschutz Medical Campus' [BIOS 6618 Recitation](/recitation/index.qmd) collection. To view other questions, you can view the [BIOS 6618 Recitation](/recitation/index.qmd) collection page or use the search bar to look for keywords.

# Population Moments

In statistics we are often interested in estimating the average of some random variable and a summary of its variability. Oftentimes, these are represented by the mean and variance (or standard deviation). However, there are other measures that can describe the shape of a distribution, such as *skewness* and *kurtosis*.

Fortunately, these all exist under a unifying framework of the **moments** of a function. However, to be more useful we may modify the *raw* moments to be either *central* moments (e.g., variance) or *standardized* moments (e.g., skewness and kurtosis).

Let's see a summary of these measures before diving a little deeper into skewness and kurtosis:

| **Moment** | **Summary** | **Type of Moment** | **Formula** |   **General Description**   |
|:----------:|:-----------:|:------------------:|:-----------:|:---------------------------:|
|     1st    |     Mean    |         Raw        | $\mu = E(X)$            |   Location of distribution  |
|     2nd    |   Variance  |       Central      | $\sigma^2 = E[(X-\mu)^2]$            | Variability of distribution |
|     3rd    |   Skewness  |    Standardized    | $\gamma_1 = E\left[\left( \frac{X-\mu}{\sigma} \right)^3 \right]$            |   Symmetry of distribution  |
|     4th    |   Kurtosis  |    Standardized    | $\text{Kurt}[X] = E\left[\left( \frac{X-\mu}{\sigma} \right)^4 \right]$            |  Tailedness of distribution |

# Skewness

Skewness is a measure of the symmetry of a distribution. Symmetric distributions like the normal, $t$, and uniform, have no skewness. However, distributions may be skewed to the right (i.e., *positively skewed*) where there is a greater probability of observing larger values of $x$ or to the left (i.e., *negatively skewed*) where there is a greater probability of observing smaller values of $x$.

Skewness may be important when chosing the appropriate summary measure. The mean, median, and mode are all equivalent with non-skewed distributions (an exception being the uniform, where any value in the sample space could be the mode). However, the mean and median will likely not match when skewness is present.

# Kurtosis

Kurtosis is a measure of the tailedness of a probability distribution. A higher value of kurtosis leads to more extreme outliers or deviations relative to a distribution with lower kurtosis.

Oftentimes we are interested in describing the *excess kurtosis* of a distribution. To do this we have to define a gold standard, which in practice is the univariate normal distribution (whose excess kurtosis is equal to 0, also known as **mesokurtic**).

A **platykurtic** distribution has negative excess kurtosis (i.e., there is lower probability of having extreme outliers/deviations in our tails). A **leptokurtic** distribution has heavier/fatter tails than a normal distribution and a greater chance of observing more extreme values.

Some examples are presented below for each form of excess kurtosis:

```{r, class.source = 'fold-hide'}
set.seed(914) # set seed for reproducibility

meso <- rnorm(n=10000)
lept <- rt(n=10000, df=3)
plat <- runif(n=10000, min=-3, max=3)

# estimate ranges to set breaks to be consistent for all distributions
min_range <- floor( min(c(meso,lept,plat)) ) # round down to closest integer
max_range <- ceiling( max(c(meso,lept,plat)) ) # round up to closest integer

par(mfrow=c(3,1)) # panel figure with 3 rows, 1 column
hist(meso, main='Mesokurtic (Normal)', xlab='x', ylab='Density', xlim=c(-4,4), breaks=seq(min_range,max_range,by=0.2), prob=T)
hist(lept, main='Leptokurtic (t)', xlab='x', ylab='Density', xlim=c(-4,4), breaks=seq(min_range,max_range,by=0.2), prob=T)
hist(plat, main='Platykurtic (Uniform)', xlab='x', ylab='Density', xlim=c(-4,4), breaks=seq(min_range,max_range,by=0.2), prob=T)
```

